\pkg{cv}: An \proglang{R} Package for Cross-Validating Regression Models

### 2024-05-21 scratch/notes_john/JSS-article/JSS-article-3.tex
add and begin to work on JSS-article-3

\author{
John Fox~\orcidlink{0000-0002-1196-8012}\\McMaster
University \And Georges Monette~\orcidlink{0000-0003-0076-5532}\\York
University
}
\title{\pkg{cv}: An \proglang{R} Package for Cross-Validation of
Regression Models}

\Plainauthor{John Fox, Georges Monette}
\Plaintitle{cv: An R Package for Cross-Validation of Regression Models}
\Shorttitle{\pkg{cv}: Cross-Validation of Regression Models}

@@

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

@@

Validating research by replication on independently collected data is a
common scientific norm. Emulating this process in a single study by
data-division is less common: The data are randomly divided into two,
possibly equal-size, parts; the first part is used to develop and fit a
statistical model; and then the second part is used to assess the
adequacy of the model fit to the first part of the data. Data-division,
however, suffers from two problems: (1) Dividing the data decreases the
sample size and thus increases sampling error; and (2), even more
disconcertingly, the results can vary substantially based on the random
division of the data, particularly in smaller samples: See \citet[Sec.
5.3]{Harrell:2015} for this and other remarks about data-division and
cross-validation.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> library("cv", quietly=TRUE)
R> methods("cv")
\end{CodeInput}
\begin{CodeOutput}
[1] cv.default* cv.glm*     cv.glmmTMB* cv.lm*      cv.lme*     cv.merMod* 
[7] cv.modList* cv.rlm*    
see '?methods' for accessing help and source code
\end{CodeOutput}
\end{CodeChunk}

@@

In the interest of brevity, we won't describe all of the features of the
\pkg{cv} package here, concentrating on the aspects of the package that
are relatively novel. For example, the \code{cv()} and \code{cvSelect()}
functions can perform computations in parallel and can independently
replicate a cross-validation procedure several times. These and other
features not discussed in this paper are taken up in the vignettes
distributed with the package, which also provides greater detail on some
of topics that we do describe, such as extensions to the package.

\hypertarget{preliminary-example-polynomial-regression}{%
\section{Preliminary Example: Polynomial
regression}\label{preliminary-example-polynomial-regression}}

@@

{\centering \includegraphics[width=0.5\linewidth]{JSS-article-reduced-2_files/figure-latex/mpg-horsepower-scatterplot-polynomials-1}

@@

\caption[mpg vs horsepower for the Auto data, showing fitted polynomials of degree 1 through 5]{mpg vs horsepower for the Auto data, showing fitted polynomials of degree 1 through 5.}\label{fig:mpg-horsepower-scatterplot-polynomials}
\end{figure}
\end{CodeChunk}

We'll focus here on the relationship of \code{mpg} (miles per gallon) to
\code{horsepower}, as displayed in Figure
\ref{fig:mpg-horsepower-scatterplot-polynomials}. The relationship
between the two variables is monotone, decreasing, and nonlinear.
Following \citet{JamesEtAl:2021}, we'll consider approximating the
relationship by a polynomial regression, with the degree of the
polynomial \(p\) ranging from 1 (a linear regression) to 10.\footnote{Although
  it serves to illustrate the use of CV, a polynomial is not the best
  choice here. Consider, for example the scatterplot for log-transformed
  \texttt{mpg} and \texttt{horsepower}, produced by
  \texttt{plot(mpg\ \textasciitilde{}\ horsepower,\ data=Auto,\ log="xy")}
  (execution of which is left to the reader). We revisit the \code{Auto}
  data in Section \ref{cross-validating-model-selection}.} Polynomial
fits for \(p = 1\) to \(5\) are shown in Figure
\ref{fig:mpg-horsepower-scatterplot-polynomials}. The linear fit is
clearly inappropriate; the fits for \(p = 2\) (quadratic) through \(4\)
are very similar; and the fit for \(p = 5\) may over-fit the data by
chasing one or two relatively high \texttt{mpg} values at the right (but
see the CV results reported below).

@@

{\centering \includegraphics[width=0.5\linewidth]{JSS-article-reduced-2_files/figure-latex/mpg-horsepower-MSE-se-1}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> args(cv:::cv.lm)
\end{CodeInput}
\begin{CodeOutput}
function (model, data = insight::get_data(model), criterion = mse, 
    k = 10, reps = 1, seed, confint = n >= 400, level = 0.95, 
    method = c("auto", "hatvalues", "Woodbury", "naive"), ncores = 1, 
    ...) 
NULL
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{itemize}
\item
  \code{model}, an \code{"lm"} object, the only required argument.
\item
  \code{data}, which can usually be inferred from the \code{model}
  object.
\item
  \code{criterion}, a function to compute the CV criterion (defaulting
  to \code{mse}).
\item
  \code{k}, the number of folds to employ (defaulting to \code{10}); the
  character value \code{"n"} or \code{"loo"} may be supplied to specify
  leave-one-out cross-validation.
\item
  \code{reps}, the number of times to repeat the CV procedure
  (defaulting to 1).
\item
  \code{seed}, the seed for \proglang{R}'s pseudo-random number
  generator; if not specified a value is randomly selected, reported,
  and saved, so that the CV procedure is replicable.
\item
  \code{confint}, whether or not to compute a confidence interval for
  the CV criterion, defaulting to \code{TRUE} if there are at least 400
  cases; a confidence interval is computed only if the CV criterion can
  be expressed as the average of casewise components (see Section
  \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}
  for details).
\item
  \code{level}, the level for the confidence interval (defaulting to
  \code{0.95}).
\item
  \code{method}, the computational method to employ: \code{"hatvalues"}
  is relevant only for LOO CV and bases computation on the hatvalues for
  the linear model; \code{"Woodbury"} employs the Woodbury matrix to
  compute the CV criterion with each fold deleted; \code{"naive"}
  updates the model using the \code{update()} function; and
  \code{"auto"} selects \code{"hatvalues"} for LOO CV and
  \code{"Woodbury"} for \(k\)-fold CV, both of which are much more
  efficient than literally updating the least-squares fit (see below and
  Section
  \ref{efficient-computations-for-linear-and-generalized-linear-models}).
\item
  \code{ncores}, the number of cores to employ for parallel computation;
  if \code{cores = 1} (the default), the computations are not
  parallelized.
\end{itemize}

@@

Residual standard error: 4.374 on 389 degrees of freedom
Multiple R-squared:  0.6876,    Adjusted R-squared:  0.686 
F-statistic:   428 on 2 and 389 DF,  p-value: < 2.2e-16
\end{CodeOutput}
\begin{CodeInput}
R> cv(m.auto, confint=TRUE)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 610656
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation
method: Woodbury
criterion: mse
cross-validation criterion = 19.39106
bias-adjusted cross-validation criterion = 19.36966
95% CI for bias-adjusted CV criterion = (15.86196, 22.87735)
full-sample criterion = 18.98477 
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> cv(m.auto, k="loo", method="naive", confint=TRUE)
\end{CodeInput}
\begin{CodeOutput}
n-Fold Cross Validation
criterion: mse
cross-validation criterion = 19.24821
bias-adjusted cross-validation criterion = 19.24787
95% CI for bias-adjusted CV criterion = (15.77884, 22.71691)
full-sample criterion = 18.98477 
\end{CodeOutput}
\begin{CodeInput}
R> cv(m.auto, k="loo", method="Woodbury", confint=TRUE)
\end{CodeInput}
\begin{CodeOutput}
n-Fold Cross Validation
method: Woodbury
criterion: mse
cross-validation criterion = 19.24821
bias-adjusted cross-validation criterion = 19.24787
95% CI for bias-adjusted CV criterion = (15.77884, 22.71691)
full-sample criterion = 18.98477 
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> m.auto.glm <- glm(mpg ~ poly(horsepower, 2), data=Auto)
R> boot::cv.glm(Auto, m.auto.glm)$delta
\end{CodeInput}
\begin{CodeOutput}
[1] 19.24821 19.24787
\end{CodeOutput}
\begin{CodeInput}
R> microbenchmark::microbenchmark(
+   hatvalues = cv(m.auto, k="loo"),
+   Woodbury = cv(m.auto, k="loo", method="Woodbury"),
+   naive = cv(m.auto, k="loo", method="naive"),
+   cv.glm = boot::cv.glm(Auto, m.auto.glm),
+   times=10
+ )
\end{CodeInput}
\begin{CodeOutput}
Warning in microbenchmark::microbenchmark(hatvalues = cv(m.auto, k = "loo"), :
less accurate nanosecond times to avoid potential integer overflows
\end{CodeOutput}
\begin{CodeOutput}
Unit: microseconds
      expr        min         lq       mean     median         uq        max
 hatvalues    981.745   1172.477   1163.211   1189.943   1203.063   1325.817
  Woodbury   9958.777  10178.496  10309.593  10232.370  10356.067  10998.127
     naive 214083.345 217623.654 247100.940 220301.795 277217.933 331491.314
    cv.glm 381699.053 383011.832 401940.298 387351.887 439080.439 442256.709
 neval cld
    10 a  
    10 a  
    10  b 
    10   c
\end{CodeOutput}
\end{CodeChunk}

On our computer, using the hatvalues is about an order of magnitude
faster than employing Woodbury matrix updates, and more than two orders
of magnitude faster than refitting the model.\footnote{Out of
  impatience, we asked \code{microbenchmark()} to execute each command
  only 10 times rather than the default 100. With the exception of the
  last column, the output is self-explanatory. The last column shows
  which methods have average timings that are statistically
  distinguishable. Because of the small number of repetitions (i.e.,
  10), the \code{"hatvalues"} and \code{"Woodbury"} methods aren't
  distinguishable, but the difference between these methods persists
  when we perform more repetitions---we invite the reader to redo this
  computation with the default \code{times=100} repetitions.}

\hypertarget{comparing-competing-models}{%
\subsection{Comparing competing
models}\label{comparing-competing-models}}

@@

{\centering \subfloat[\label{fig:polynomial-regression-CV-graph-2-1}]{\includegraphics[width=0.5\linewidth]{JSS-article-reduced-2_files/figure-latex/polynomial-regression-CV-graph-2-1} }\subfloat[\label{fig:polynomial-regression-CV-graph-2-2}]{\includegraphics[width=0.5\linewidth]{JSS-article-reduced-2_files/figure-latex/polynomial-regression-CV-graph-2-2} }

@@

\hypertarget{cross-validating-mixed-effects-models}{%
\section{Cross-validating mixed-effects
models}\label{cross-validating-mixed-effects-models}}

@@

In contrast, mixed-effects models are fit to \emph{dependent} data, in
which cases are clustered, such as hierarchical data, where the clusters
comprise higher-level units (e.g., students clustered in schools), or
longitudinal data, where the clusters are individuals and the cases are
repeated observations on the individuals over time.\footnote{There are,
  however, more complex situations that give rise to so-called
  \emph{crossed} (rather than \emph{nested}) random effects. For
  example, consider students within classes within schools. In primary
  schools, students typically are in a single class, and so classes are
  nested within schools. In secondary schools, however, students
  typically take several classes and students who are together in a
  particular class may not be together in other classes; consequently,
  random effects based on classes within schools are crossed. The
  \code{lmer()} function in the \pkg{lme4} package is capable of
  modeling both nested and crossed random effects, and the \code{cv()}
  methods for mixed models in the \pkg{cv} package pertain to both
  nested and crossed random effects. We present an example of the latter
  in a vignette for the \pkg{cv} package.}

@@

\hypertarget{example-the-high-school-and-beyond-data}{%
\subsection{Example: The High-School and Beyond
data}\label{example-the-high-school-and-beyond-data}}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> cv(hsb.lmer, k=10, clusterVariables="school", seed=5240)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 5240
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation based on 160 {school} clusters
cross-validation criterion = 39.15662
bias-adjusted cross-validation criterion = 39.14844
95% CI for bias-adjusted CV criterion = (38.06554, 40.23135)
full-sample criterion = 39.00599 
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> cv(hsb.lmer, seed=1575)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 1575
\end{CodeOutput}
\begin{CodeOutput}
Warning in checkConv(attr(opt, "derivs"), opt$par, ctrl = control$checkConv, :
Model failed to converge with max|grad| = 0.00587228 (tol = 0.002, component 1)
\end{CodeOutput}
\begin{CodeOutput}
boundary (singular) fit: see help('isSingular')
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation
cross-validation criterion = 37.44473
bias-adjusted cross-validation criterion = 37.33801
95% CI for bias-adjusted CV criterion = (36.28761, 38.38841)
full-sample criterion = 36.06767 
\end{CodeOutput}
\end{CodeChunk}

@@

\hypertarget{example-contrasting-cluster-based-and-case-based-cv}{%
\subsection{Example: Contrasting cluster-based and case-based
CV}\label{example-contrasting-cluster-based-and-case-based-cv}}

@@

{\centering \subfloat[\label{fig:plot1-1}]{\includegraphics[width=0.45\linewidth]{JSS-article-reduced-2_files/figure-latex/plot1-1} }\subfloat[\label{fig:plot1-2}]{\includegraphics[width=0.45\linewidth]{JSS-article-reduced-2_files/figure-latex/plot1-2} }\newline\subfloat[\label{fig:plot1-3}]{\includegraphics[width=0.45\linewidth]{JSS-article-reduced-2_files/figure-latex/plot1-3} }\subfloat[\label{fig:plot1-4}]{\includegraphics[width=0.45\linewidth]{JSS-article-reduced-2_files/figure-latex/plot1-4} }\newline\subfloat[\label{fig:plot1-5}]{\includegraphics[width=0.45\linewidth]{JSS-article-reduced-2_files/figure-latex/plot1-5} }

@@

{\centering \subfloat[\label{fig:cross-validation-clusters-1}]{\includegraphics[width=0.45\linewidth]{JSS-article-reduced-2_files/figure-latex/cross-validation-clusters-1} }\subfloat[\label{fig:cross-validation-clusters-2}]{\includegraphics[width=0.45\linewidth]{JSS-article-reduced-2_files/figure-latex/cross-validation-clusters-2} }

@@

\hypertarget{cross-validating-model-specification}{%
\section{Cross-validating model
specification}\label{cross-validating-model-specification}}

@@

The \code{cvSelect()} function in the \pkg{cv} package cross-validates a
model-specification process in a general manner. Functions for three
such model-specification processes are included in the package:
\code{selectStepAIC()}, based on the \code{stepAIC()} function in the
\pkg{MASS} package \citep{VenablesRipley:2002}, performs stepwise
predictor selection for regression models; \code{selectTrans()}, based
on the \code{powerTransform()} function in the \pkg{car} package
\citep{FoxWeisberg:2019}, transforms predictors and the response in a
regression model towards normality; and
\code{selectTransAndStepAIC()}---the use of which we illustrate in the
current section---performs both of these procedures sequentially. In the
vignette on extending the \pkg{cv} package, we explain how to add
model-selection procedures.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> Auto$cylinders <- factor(Auto$cylinders,
+                          labels=c("3.4", "3.4", "5.6", "5.6", "8"))
R> Auto$year <- as.factor(Auto$year)
R> Auto$origin <- factor(Auto$origin,
+                       labels=c("America", "Europe", "Japan"))
R> rownames(Auto) <- make.names(Auto$name, unique=TRUE)
R> Auto$name <- NULL
R> 
R> scatterplotMatrix(~ mpg + displacement + horsepower + weight + acceleration, 
+                   smooth=list(spread=FALSE), data=Auto, pch=".")
\end{CodeInput}
\begin{figure}

{\centering \includegraphics[width=0.6\linewidth]{JSS-article-reduced-2_files/figure-latex/Auto-explore-1}

@@

\caption[Scatterplot matrix for the numeric variables in the Auto data]{Scatterplot matrix for the numeric variables in the Auto data}\label{fig:Auto-explore}
\end{figure}
\end{CodeChunk}

A comment before we proceed: \code{origin} is clearly categorical and so
converting it to a factor is natural, but we could imagine treating
\code{cylinders} and \code{year} as numeric predictors. There are,
however, only 5 distinct values of \code{cylinders} (ranging from 3 to
8), but cars with 3 or 5 cylinders are rare, and none of the cars has 7
cylinders. There are similarly only 13 distinct years between 1970 and
1982 in the data, and the relationship between \code{mpg} and
\code{year} is difficult to characterize.\footnote{Making the decision
  to treat \code{year} as a factor on this basis could be construed as
  cheating in the current context, which illustrates the difficulty of
  automating the whole model-selection process. It's rarely desirable,
  in our opinion, to forgo exploration of the data to ensure the purity
  of model validation. We believe, however, that it's still useful to
  automate as much of the process as we can to obtain a more realistic,
  if still biased, estimate of the predictive power of a model.} It's
apparent that most these variables are positively skewed and that many
of the pairwise relationships among them are nonlinear.

@@

{\centering \includegraphics[width=0.6\linewidth]{JSS-article-reduced-2_files/figure-latex/Auto-working-model-1}

@@

\caption[Component+residual plots for the working model fit to the Auto data]{Component+residual plots for the working model fit to the Auto data}\label{fig:Auto-working-model}
\end{figure}
\end{CodeChunk}

@@

{\centering \includegraphics[width=0.6\linewidth]{JSS-article-reduced-2_files/figure-latex/Auto-transformed-scatterplot-matrix-1}

@@

{\centering \includegraphics[width=0.6\linewidth]{JSS-article-reduced-2_files/figure-latex/Auto-CR-plots-transformed-1}

>>>>

\author{
John Fox~\orcidlink{0000-0002-1196-8012}\\McMaster
University \And Georges Monette~\orcidlink{0000-0003-0076-5532}\\York
University
}
\title{\pkg{cv}: An \proglang{R} Package for Cross-Validating Regression
Models}

\Plainauthor{John Fox, Georges Monette}
\Plaintitle{cv: An R Package for Cross-Validating Regression Models}
\Shorttitle{\pkg{cv}: Cross-Validating Regression Models}

@@

\section{Introduction}\label{introduction}

@@

Validating research by replication on independently collected data is a
common scientific norm. Emulating this process in a single study by
data-division is less common: The data are randomly divided into two,
possibly equal-size, parts; the first part is used to develop and fit a
statistical model; and then the second part is used to assess the
adequacy of the model fit to the first part of the data. Data-division,
however, suffers from two problems: (1) Dividing the data decreases the
sample size and thus increases sampling error; and (2), even more
disconcertingly, the results can vary substantially based on the random
division of the data, particularly in smaller samples. See \citet[Sec.
5.3]{Harrell:2015} for this and other remarks about data-division and
cross-validation.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> library("cv", quietly=TRUE)
R> methods("cv")
\end{CodeInput}
\begin{CodeOutput}
[1] cv.default*  cv.function* cv.glm*      cv.glmmTMB*  cv.lm*      
[6] cv.lme*      cv.merMod*   cv.modList*  cv.rlm*     
see '?methods' for accessing help and source code
\end{CodeOutput}
\end{CodeChunk}

@@

A number of existing \proglang{R} packages include functions for
cross-validating regression models. We briefly situate the \pkg{cv}
package relative to other \proglang{R} software for cross-validation in
Section \ref{comparing-cv-to-other-r-software-for-cross-validation}.

In the interest of brevity, we won't describe all of the features of the
\pkg{cv} package here, concentrating on the aspects of the package that
are relatively novel. For example, the \code{cv()} function can perform
computations in parallel and can independently replicate a
cross-validation procedure several times. These and other features not
discussed in this paper are taken up in the vignettes distributed with
the package, which also provides greater detail on some of topics that
we do describe, such as extensions to the package.

\section{Preliminary Example: Polynomial
regression}\label{preliminary-example-polynomial-regression}

@@

{\centering \includegraphics[width=0.5\linewidth]{JSS-article-3_files/figure-latex/mpg-horsepower-scatterplot-polynomials-1}

@@

\caption{\code{mpg} vs. \code{horsepower} for the \code{Auto} data, showing fitted polynomials of degree 1 through 5.}\label{fig:mpg-horsepower-scatterplot-polynomials}
\end{figure}
\end{CodeChunk}

We'll focus here on the relationship of \code{mpg} (miles per gallon) to
\code{horsepower}, as displayed in Figure
\ref{fig:mpg-horsepower-scatterplot-polynomials}. The relationship
between the two variables is monotone, decreasing, and nonlinear.
Following \citet{JamesEtAl:2021}, we'll consider approximating the
relationship by a polynomial regression, with the degree of the
polynomial \(p\) ranging from 1 (a linear regression) to 10.\footnote{Although
  it serves to illustrate the use of CV, a polynomial is not the best
  choice here. Consider, for example the scatterplot for log-transformed
  \texttt{mpg} and \texttt{horsepower}, produced by
  \texttt{plot(mpg\ \textasciitilde{}\ horsepower,\ data=Auto,\ log="xy")}
  (execution of which is left to the reader). We revisit the \code{Auto}
  data in Section \ref{cross-validating-model-specification}.}
Polynomial fits for \(p = 1\) to \(5\) are shown in Figure
\ref{fig:mpg-horsepower-scatterplot-polynomials}. The linear fit is
clearly inappropriate; the fits for \(p = 2\) (quadratic) through \(4\)
are very similar; and the fit for \(p = 5\) may over-fit the data by
chasing one or two relatively high \texttt{mpg} values at the right (but
see the CV results reported below).

@@

{\centering \includegraphics[width=0.5\linewidth]{JSS-article-3_files/figure-latex/mpg-horsepower-MSE-se-1}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> args(cv:::cv.lm)
\end{CodeInput}
\begin{CodeOutput}
function (model, data = insight::get_data(model), criterion = mse, 
    k = 10L, reps = 1L, seed = NULL, details = k <= 10L, confint = n >= 
        400L, level = 0.95, method = c("auto", "hatvalues", "Woodbury", 
        "naive"), ncores = 1L, ...) 
NULL
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{itemize}
\item
  \code{model}, an \code{"lm"} object, the only required argument.
\item
  \code{data}, which can usually be inferred from the \code{model}
  object.
\item
  \code{criterion}, a function to compute the CV criterion (defaulting
  to \code{mse}).
\item
  \code{k}, the number of folds to employ (defaulting to \code{10}); the
  character value \code{"n"} or \code{"loo"} may be supplied to specify
  leave-one-out cross-validation.
\item
  \code{reps}, the number of times to repeat the CV procedure
  (defaulting to 1).
\item
  \code{seed}, the seed for \proglang{R}'s pseudo-random number
  generator; if not specified a value is randomly selected, reported,
  and saved, so that the CV procedure is replicable.
\item
  \code{confint}, whether or not to compute a confidence interval for
  the CV criterion, defaulting to \code{TRUE} if there are at least 400
  cases; a confidence interval is computed only if the CV criterion can
  be expressed as the average of casewise components (see Section
  \ref{computation-of-the-bias-corrected-cv-criterion-and-confidence-intervals}
  for details).
\item
  \code{level}, the level for the confidence interval (defaulting to
  \code{0.95}).
\item
  \code{method}, the computational method to employ: \code{"hatvalues"}
  is relevant only for LOO CV and bases computation on the hatvalues for
  the linear model; \code{"Woodbury"} employs the Woodbury matrix
  identity to compute the CV criterion with each fold deleted;
  \code{"naive"} updates the model using the \code{update()} function;
  and \code{"auto"} selects \code{"hatvalues"} for LOO CV and
  \code{"Woodbury"} for \(k\)-fold CV, both of which are much more
  efficient than literally updating the least-squares fit (see below and
  Section
  \ref{efficient-computations-for-linear-and-generalized-linear-models}).
\item
  \code{ncores}, the number of cores to employ for parallel computation;
  if \code{cores = 1} (the default), the computations are not
  parallelized.
\end{itemize}

@@

Residual standard error: 4.374 on 389 degrees of freedom
Multiple R-squared:  0.6876,    Adjusted R-squared:  0.686 
F-statistic:   428 on 2 and 389 DF,  p-value: < 2.2e-16
\end{CodeOutput}
\begin{CodeInput}
R> cv(m.auto, confint=TRUE)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 32663
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation
method: Woodbury
criterion: mse
cross-validation criterion = 19.24054
bias-adjusted cross-validation criterion = 19.22691
95% CI for bias-adjusted CV criterion = (15.76529, 22.68854)
full-sample criterion = 18.98477 
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> cv(m.auto, k="loo", method="naive", confint=TRUE)
\end{CodeInput}
\begin{CodeOutput}
n-Fold Cross Validation
method: naive
criterion: mse
cross-validation criterion = 19.24821
bias-adjusted cross-validation criterion = 19.24787
95% CI for bias-adjusted CV criterion = (15.77884, 22.71691)
full-sample criterion = 18.98477 
\end{CodeOutput}
\begin{CodeInput}
R> cv(m.auto, k="loo", method="Woodbury", confint=TRUE)
\end{CodeInput}
\begin{CodeOutput}
n-Fold Cross Validation
method: Woodbury
criterion: mse
cross-validation criterion = 19.24821
bias-adjusted cross-validation criterion = 19.24787
95% CI for bias-adjusted CV criterion = (15.77884, 22.71691)
full-sample criterion = 18.98477 
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> m.auto.glm <- glm(mpg ~ poly(horsepower, 2), data=Auto)
R> boot::cv.glm(Auto, m.auto.glm)$delta
\end{CodeInput}
\begin{CodeOutput}
[1] 19.24821 19.24787
\end{CodeOutput}
\begin{CodeInput}
R> microbenchmark::microbenchmark(
+   hatvalues = cv(m.auto, k="loo"),
+   Woodbury = cv(m.auto, k="loo", method="Woodbury"),
+   naive = cv(m.auto, k="loo", method="naive"),
+   cv.glm = boot::cv.glm(Auto, m.auto.glm),
+   times=10
+ )
\end{CodeInput}
\begin{CodeOutput}
Warning in microbenchmark::microbenchmark(hatvalues = cv(m.auto, k = "loo"), :
less accurate nanosecond times to avoid potential integer overflows
\end{CodeOutput}
\begin{CodeOutput}
Unit: milliseconds
      expr        min         lq       mean     median         uq        max
 hatvalues   1.015119   1.201546   1.191919   1.206446   1.222333   1.258987
  Woodbury  12.108407  12.390569  19.157808  12.634949  15.289884  73.319275
     naive 219.469269 220.993977 223.441857 223.066916 226.155180 228.677377
    cv.glm 379.740606 383.517075 390.172306 384.790925 386.413397 442.643421
 neval
    10
    10
    10
    10
\end{CodeOutput}
\end{CodeChunk}

On our computer, using the hatvalues is about an order of magnitude
faster than employing Woodbury matrix updates, and more than two orders
of magnitude faster than refitting the model.

\subsection{Comparing competing
models}\label{comparing-competing-models}

@@

{\centering \subfloat[\label{fig:polynomial-regression-CV-graph-2-1}]{\includegraphics[width=0.5\linewidth]{JSS-article-3_files/figure-latex/polynomial-regression-CV-graph-2-1} }\subfloat[\label{fig:polynomial-regression-CV-graph-2-2}]{\includegraphics[width=0.5\linewidth]{JSS-article-3_files/figure-latex/polynomial-regression-CV-graph-2-2} }

@@

\section{Cross-validating mixed-effects
models}\label{cross-validating-mixed-effects-models}

@@

In contrast, mixed-effects models are fit to \emph{dependent} data, in
which cases are clustered, such as hierarchical data, where the clusters
comprise higher-level units (e.g., students clustered in schools), or
longitudinal data, where the clusters are individuals and the cases are
repeated observations on the individuals over time.\footnote{There are,
  however, more complex situations that give rise to so-called
  \emph{crossed} (rather than \emph{nested}) random effects. For
  example, consider students within classes within schools. In primary
  schools, students typically are in a single class, and so classes are
  nested within schools. In secondary schools, however, students
  typically take several classes and students who are together in a
  particular class may not be together in other classes; consequently,
  random effects based on classes within schools are crossed. The
  \code{lmer()} function in the \pkg{lme4} package, for example, is
  capable of modeling both nested and crossed random effects, and the
  \code{cv()} methods for mixed models in the \pkg{cv} package pertain
  to both nested and crossed random effects. We present an example of
  the latter in a vignette for the \pkg{cv} package.}

@@

\subsection{Example: The High-School and Beyond
data}\label{example-the-high-school-and-beyond-data}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> cv(hsb.lmer, k=10, clusterVariables="school", seed=5240)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 5240
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation based on 160 {school} clusters
criterion: mse
cross-validation criterion = 39.15662
bias-adjusted cross-validation criterion = 39.14844
95% CI for bias-adjusted CV criterion = (38.06554, 40.23135)
full-sample criterion = 39.00599 
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> cv(hsb.lmer, seed=1575)
\end{CodeInput}
\begin{CodeOutput}
R RNG seed set to 1575
\end{CodeOutput}
\begin{CodeOutput}
Warning in checkConv(attr(opt, "derivs"), opt$par, ctrl = control$checkConv, :
Model failed to converge with max|grad| = 0.00587207 (tol = 0.002, component 1)
\end{CodeOutput}
\begin{CodeOutput}
boundary (singular) fit: see help('isSingular')
\end{CodeOutput}
\begin{CodeOutput}
10-Fold Cross Validation
criterion: mse
cross-validation criterion = 37.44473
bias-adjusted cross-validation criterion = 37.33801
95% CI for bias-adjusted CV criterion = (36.28761, 38.38841)
full-sample criterion = 36.06767 
\end{CodeOutput}
\end{CodeChunk}

@@

\subsection{Example: Contrasting cluster-based and case-based
CV}\label{example-contrasting-cluster-based-and-case-based-cv}

@@

{\centering \subfloat[\label{fig:plot1-1}]{\includegraphics[width=0.45\linewidth]{JSS-article-3_files/figure-latex/plot1-1} }\subfloat[\label{fig:plot1-2}]{\includegraphics[width=0.45\linewidth]{JSS-article-3_files/figure-latex/plot1-2} }\newline\subfloat[\label{fig:plot1-3}]{\includegraphics[width=0.45\linewidth]{JSS-article-3_files/figure-latex/plot1-3} }\subfloat[\label{fig:plot1-4}]{\includegraphics[width=0.45\linewidth]{JSS-article-3_files/figure-latex/plot1-4} }\newline\subfloat[\label{fig:plot1-5}]{\includegraphics[width=0.45\linewidth]{JSS-article-3_files/figure-latex/plot1-5} }

@@

{\centering \subfloat[\label{fig:cross-validation-clusters-1}]{\includegraphics[width=0.45\linewidth]{JSS-article-3_files/figure-latex/cross-validation-clusters-1} }\subfloat[\label{fig:cross-validation-clusters-2}]{\includegraphics[width=0.45\linewidth]{JSS-article-3_files/figure-latex/cross-validation-clusters-2} }

@@

\section{Cross-validating model
specification}\label{cross-validating-model-specification}

@@

The \code{"function"} method for \code{cv()} cross-validates a
model-specification process in a general manner. Functions for four such
model-specification processes are included in the package:
\code{selectStepAIC()}, based on the \code{stepAIC()} function in the
\pkg{MASS} package \citep{VenablesRipley:2002}, performs stepwise
predictor selection for regression models; \code{selectTrans()}, based
on the \code{powerTransform()} function in the \pkg{car} package
\citep{FoxWeisberg:2019}, transforms predictors and the response in a
regression model towards normality; \code{selectTransAndStepAIC()}---the
use of which we illustrate in the current section---performs both of
these procedures sequentially; and \code{selectModelList()}---also
illustrated in the current section---uses CV both to select one of
several competing models, and then, recursively, to estimate prediction
error for the selected model. In a vignette on extending the \pkg{cv}
package, we explain how to add model-selection procedures.

\subsection{Example: Data transformation and predictor selection for the
Auto
data}\label{example-data-transformation-and-predictor-selection-for-the-auto-data}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> Auto$cylinders <- factor(Auto$cylinders,
+                          labels=c("3-4", "3-4", "5-6", "5-6", "8"))
R> Auto$year <- as.factor(Auto$year)
R> Auto$origin <- factor(Auto$origin,
+                       labels=c("America", "Europe", "Japan"))
R> rownames(Auto) <- make.names(Auto$name, unique=TRUE)
R> Auto$name <- NULL
R> 
R> scatterplotMatrix(~ mpg + displacement + horsepower + weight + acceleration, 
+                   smooth=list(spread=FALSE), data=Auto, pch=".")
\end{CodeInput}
\begin{figure}

{\centering \includegraphics[width=0.6\linewidth]{JSS-article-3_files/figure-latex/Auto-explore-1}

@@

\caption[Scatterplot matrix for the numeric variables in the \code{Auto} data]{Scatterplot matrix for the numeric variables in the \code{Auto} data}\label{fig:Auto-explore}
\end{figure}
\end{CodeChunk}

A comment before we proceed: \code{origin} is clearly categorical and so
converting it to a factor is natural, but we could imagine treating
\code{cylinders} and \code{year} as numeric predictors. There are,
however, only 5 distinct values of \code{cylinders} (ranging from 3 to
8), but cars with 3 or 5 cylinders are rare, and none of the cars has 7
cylinders. There are similarly only 13 distinct years between 1970 and
1982 in the data, and the relationship between \code{mpg} and
\code{year} is difficult to characterize.\footnote{Making the decision
  to treat \code{year} as a factor on this basis could be construed as
  cheating in the current context, which illustrates the difficulty of
  automating the whole model-selection process. It's rarely desirable,
  in our opinion, to forgo exploration of the data to ensure the purity
  of model validation. We believe, however, that it's still useful to
  automate as much of the process as we can, as illustrated here, to
  obtain a more realistic, if still biased, estimate of the predictive
  power of a model.} It's apparent that most these variables are
positively skewed and that many of the pairwise relationships among them
are nonlinear.

@@

{\centering \includegraphics[width=0.6\linewidth]{JSS-article-3_files/figure-latex/Auto-working-model-1}

@@

\caption[Component+residual plots for the working model fit to the \code{Auto} data]{Component+residual plots for the working model fit to the \code{Auto} data}\label{fig:Auto-working-model}
\end{figure}
\end{CodeChunk}

@@

{\centering \includegraphics[width=0.6\linewidth]{JSS-article-3_files/figure-latex/Auto-transformed-scatterplot-matrix-1}

@@

{\centering \includegraphics[width=0.6\linewidth]{JSS-article-3_files/figure-latex/Auto-CR-plots-transformed-1}
